---
title: "Data Cleaning for MLB EDA"
author: "Jonah Lorenzo, Nicolas Delfau, Asher Heflin"
date: "March 2026"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("ggh4x")
```

```{r}
library(tidyverse)
library(knitr)
library(readr)
library(ggrepel)
library(scales)
library(ggh4x)
```

Load Data:
```{r}
statcast <- read_csv("statcast_2017.csv") %>%
  rbind(read_csv("statcast_2018.csv")) %>%
  rbind(read_csv("statcast_2019.csv")) %>%
  rbind(read_csv("statcast_2020.csv")) %>%
  rbind(read_csv("statcast_2021.csv")) %>%
  rbind(read_csv("statcast_2022.csv")) %>%
  rbind(read_csv("statcast_2023.csv")) %>%
  rbind(read_csv("statcast_2024.csv")) %>%
  rbind(read_csv("statcast_2025.csv"))
```

Save Combined Data:
```{r}
write_csv(statcast, "statcast_full.csv")
```

Read Combined Data:
```{r}
statcast <- read_csv("statcast_full.csv")
```

Read FanGraphs Data:
```{r}
fangraphs <- read_csv("fangraphs.csv")
```

Remove Unneeded Cols:
```{r}
statcast_clean <- statcast %>%
  select(-matches(".*deprecated$"), -spin_dir, -umpire) %>% #Remove deprecated numbers
  select(-matches("fielder_[0-9]")) %>% 
  select(-matches("on_[1-3]b$")) %>%
  select(-matches("[io]f_fielding_alignment")) %>%
  select(-matches("delta_.*")) %>%
  select(-matches("intercept_ball_.*")) %>%
  select(-matches(".*_days_since_.*"), -matches(".*_days_until_.*")) %>%
  select(-age_pit_legacy, -age_bat_legacy) %>%
  select(-matches("api_.*")) %>%
  select(-woba_value, -woba_denom) %>%
  select(-matches(".*score.*"))
  
statcast_clean <- statcast_clean %>%
  drop_na(pitch_type) %>%
  filter(game_type=="R")
#write_csv(statcast_clean, "statcast_clean.csv")
```

Format Columns Where Needed:
```{r}
statcast_clean <- statcast_clean %>%
  mutate(pitch_type = as.factor(pitch_type)) %>%
  mutate(description = as.factor(description)) %>%
  mutate(zone = as.factor(zone)) %>%
  mutate(p_throws = as.factor(p_throws))
```

Cleanup Fangraphs Data:
```{r}
fangraphs_clean <- fangraphs %>%
  select(-"...1")
```
Pitch Colors:
```{r}
pitch_levels <- statcast_clean$pitch_name %>%
  unique() %>%
  sort()

statcast_clean <- statcast_clean %>%
  mutate(pitch_name = factor(pitch_name, levels = pitch_levels))

pitch_colors <- setNames(
  hue_pal()(length(pitch_levels)),
  pitch_levels
)
```

Where pitches appear in zone: TODO
```{r}
statcast_clean %>%
  drop_na(plate_x, plate_z) %>%
  filter(pitch_type %in% c("FF", "FT", "FC", "FS", "SI", "CU", "SL", "CH", "KC")) %>%
  mutate(plate_x_in = 12 * plate_x, plate_z_in = 12 *plate_z) %>%
  ggplot(aes(x=plate_x_in, y=plate_z_in, color=pitch_type)) +
  geom_vline(xintercept = 7.5) +
  geom_vline(xintercept = -7.5) +
  geom_hline(yintercept = mean(statcast_clean$sz_bot, na.rm=TRUE)*12) +
  geom_hline(yintercept = mean(statcast_clean$sz_top, na.rm=TRUE)*12) +
  stat_ellipse(type="t", level=0.8) +
  facet_wrap(~p_throws) +
  scale_x_continuous(limits = c(-25,25),
                              breaks = seq(-25,25, 5),
                              labels = scales::number_format(suffix = "\"")) + 
  scale_y_continuous(limits = c(0,50),
                              breaks = seq(0,50, 5),
                              labels = scales::number_format(suffix = "\"")) +
  coord_equal() + 
  labs(title = "Pitch Locations At Plate",
       subtitle = "2017-2025 (ex 2020, 2022) Regular Seasons | Catcher's Perspective",
       caption = "Data: Baseball Savant via baseballr",
       x = "Distance from Center of Plate",
       y = "Height off Ground",
       color = "Pitch Type") +
  scale_color_manual(values = pitch_colors,
                     limits = pitch_levels)
```

Pitch Break:
```{r}
centers <- statcast_clean %>%
  filter(pitch_type %in% c("FF", "FT", "FC", "FS", "SI", "CU", "SL", "CH", "KC", "ST")) %>%
  group_by(pitch_type, pitch_name, p_throws) %>%
  summarise(
    x = mean(pfx_x, na.rm = TRUE)*12,
    y = mean(pfx_z, na.rm = TRUE)*12,
    .groups = "drop"
  )

statcast_clean %>%
  drop_na(pfx_x, pfx_z) %>%
  filter(pitch_type %in% c("FF", "FT", "FC", "FS", "SI", "CU", "SL", "CH", "KC", "ST")) %>%
  mutate(pfx_x_in = 12 * pfx_x, pfx_z_in = 12 * pfx_z) %>%
  ggplot(aes(x=pfx_x_in, y=pfx_z_in, color=pitch_name)) +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  stat_ellipse(type="t", linetype=2, level=0.9) +
  geom_label_repel(data = centers,
            aes(x = x, y = y, label = pitch_name, color=pitch_name),
            size=2,
            show.legend = FALSE) +
  facet_wrap(~p_throws) +
  scale_x_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  scale_y_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  coord_equal() +
  theme_bw() +
  labs(title="Induced Movement of MLB Pitches by Pitch Type",
       subtitle = "Catcher's Perspective | Grouped by Pitcher Throwing Arm",
       x="Horizontal Movement",
       y="Induced Vertical Break",
       caption="Data Source: Statcast via baseballr",
       color="Pitch Type") +
  scale_color_manual(values = pitch_colors,
                     limits = pitch_levels)
  
```


```{r}
centers_sl_st <- statcast_clean %>%
  filter(pitch_type %in% c("SL", "ST")) %>%
  group_by(pitch_type, pitch_name, p_throws) %>%
  summarise(
    x = mean(pfx_x, na.rm = TRUE)*12,
    y = mean(pfx_z, na.rm = TRUE)*12,
    .groups = "drop"
  )

statcast_clean %>%
  drop_na(pfx_x, pfx_z) %>%
  filter(pitch_type %in% c("SL", "ST")) %>%
  mutate(pfx_x_in = 12 * pfx_x, pfx_z_in = 12 * pfx_z) %>%
  ggplot(aes(x=pfx_x_in, y=pfx_z_in, color=pitch_name)) +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  stat_ellipse(type="t", linetype=2, level=0.9) +
  geom_label_repel(data = centers_sl_st,
            aes(x = x, y = y, label = pitch_name, color=pitch_name),
            size=2,
            show.legend = FALSE) +
  facet_wrap(~p_throws) +
  scale_x_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  scale_y_continuous(limits=c(-20,20),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  theme_bw() +
  labs(title="Induced Movement of Sliders and Sweepers",
       subtitle = "Catcher's Perspective | Grouped by Pitcher Throwing Arm",
       x="Horizontal Movement",
       y="Induced Vertical Break",
       caption="Data Source: Statcast via baseballr",
       color="Pitch Type") +
  scale_color_manual(values = pitch_colors,
                     limits = c("Sweeper", "Slider"))
```


```{r}
spin_and_speed <- statcast_clean %>%
  filter(pitch_type %in% c("SL", "ST")) %>%
  drop_na(release_speed, release_spin_rate) %>%
  group_by(pitch_name, game_year) %>%
  summarise(avg_spin = mean(release_spin_rate), avg_speed = mean(release_speed), .groups = "drop") %>%
  pivot_longer(cols = c(avg_speed, avg_spin), names_to="metric") %>%
  complete(pitch_name = c("Slider", "Sweeper"), metric, game_year = full_seq(game_year, 1))

mean_spin_and_speed <- spin_and_speed %>%
  group_by(metric, pitch_name) %>%
  summarise(mean_value = mean(value, na.rm=TRUE), .groups="drop")

spin_and_speed %>%
  ggplot(aes(x=game_year, y=value, color=pitch_name)) +
  geom_point() +
  geom_line() +
  geom_hline(data=mean_spin_and_speed,
            aes(yintercept=mean_value, color=pitch_name),
            linetype="dashed",
            linewidth = 1) +
  facet_wrap(~ metric, 
             scales = "free_y",
             labeller = labeller(
               metric = c(
                 avg_speed = "Average Release Speed (mph)",
                 avg_spin = "Average Spin Rate (rpm)"
               )
             )) +
  theme_bw() +
  labs(title="Spin Rate and Release Speed Comparison",
       subtitle = "Sliders vs Sweepers",
       caption="Data Source: Statcast via baseballr",
       x="Year",
       y="Speed/Spin Rate",
       color="Pitch Type") +
  scale_color_manual(values = pitch_colors,
                     limits = c("Sweeper", "Slider"))
  
```


```{r}
centers_sl_year <- statcast_clean %>%
  filter(pitch_type %in% c("FF")) %>%
  group_by(game_year, p_throws) %>%
  summarise(
    x = mean(pfx_x, na.rm = TRUE)*12,
    y = mean(pfx_z, na.rm = TRUE)*12,
    .groups = "drop"
  )

statcast_clean %>%
  drop_na(pfx_x, pfx_z) %>%
  filter(pitch_type %in% c("FF")) %>%
  mutate(pfx_x_in = 12 * pfx_x, pfx_z_in = 12 * pfx_z) %>%
  ggplot(aes(x=pfx_x_in, y=pfx_z_in, color=as.factor(home_team))) +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  stat_ellipse(type="t", linetype=2, level=0.9) +
  #geom_label(data = centers_sl_year,
  #          aes(x = x, y = y, label = as.factor(game_year), color=as.factor(game_year)),
  #          size=2,
  #          show.legend = FALSE) +
  facet_wrap(~p_throws) +
  scale_x_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  scale_y_continuous(limits=c(-20,20),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  coord_equal() +
  theme_bw()
```

Pitch Spin: TODO
```{r}
statcast_sample <- statcast_clean %>%
  slice_sample(prop=0.1)

statcast_sample %>%
  filter(pitch_type %in% c("SL","ST")) %>%
  drop_na(spin_axis) %>%
  mutate(clock_pos = spin_axis/30) %>%
  ggplot(aes(x=clock_pos, fill=pitch_type)) +
  geom_histogram(bins=12) +
  facet_wrap(~p_throws) +
  scale_color_manual(values = pitch_colors,
                     limits = c("SL", "ST")) +
  coord_polar() +
  scale_x_continuous(limits = c(1,12))
```

Pitch Mix (from Statcast Data): TODO
```{r}
pitch_props <- statcast_clean %>%
  group_by(game_year, pitch_name, pitch_type) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(game_year) %>%
  mutate(prop = n/sum(n)) %>%
  filter(pitch_type %in% c("FF", "FT", "FC", "FS", "SI", "CU", "SL", "CH", "KC", "ST"))

pitch_props %>%
  ggplot(aes(x=game_year, y=prop, fill=pitch_name)) +
  geom_area(aes(fill=pitch_name)) +
  geom_label(data = filter(pitch_props, pitch_type %in% c("ST", "SL", "SI"), game_year %in% c(2017,2019,2021,2023,2025)), aes(label=scales::percent(prop, accuracy = 0.1), color=pitch_name),
                    position = position_stack(vjust=0.5),
                    size=3,
                    fill = "white",
                    show.legend = FALSE) +
  scale_color_manual(values = pitch_colors) +
  theme_bw() +
  labs(title="Evolution of Pitch Mix",
       subtitle="Statcast Dataset",
       x="Season",
       y="Proportion",
       fill="Pitch Type") +
  scale_x_continuous(limits=c(2017,2025),
                     breaks=seq(2017,2025,1))
```

```{r}
pitcher_year <- statcast_clean %>%
  group_by(player_name, game_year, pitch_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(player_name, game_year) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

wide_pitcher <- pitcher_year %>%
  select(player_name, game_year, pitch_name, prop) %>%
  pivot_wider(names_from = pitch_name, values_from = prop, values_fill = 0)

wide_pitcher <- wide_pitcher %>%
  arrange(player_name, game_year) %>%
  group_by(player_name) %>%
  mutate(
    d_sweeper = Sweeper - lag(Sweeper),
    d_sinker  = Sinker  - lag(Sinker),
    d_slider  = Slider  - lag(Slider)) %>%
  ungroup()

cor(wide_pitcher$d_sweeper, wide_pitcher$d_sinker, use = "complete.obs")
cor(wide_pitcher$d_sweeper, wide_pitcher$d_slider, use = "complete.obs")
```

Statcast with FG:
```{r}
statcast_fg <- statcast_clean %>%
  right_join(fangraphs, by = join_by(pitcher == xMLBAMID, game_year == Season))
```

```{r}
test <- statcast_fg %>%
  filter(pitch_type=="SL") %>%
  group_by(game_year, pitcher) %>%
  summarise(
    sp_s_SL = first(sp_s_SL),
    p_throws = first(p_throws),
    n = n(),
    .groups="drop"
  ) %>%
  filter(n > 100) %>%
  group_by(game_year, p_throws) %>%
  mutate(
    tier = as.factor(case_when(
      sp_s_SL >= quantile(sp_s_SL, 0.9, na.rm = TRUE) ~ "Best",
      sp_s_SL <= quantile(sp_s_SL, 0.1, na.rm = TRUE) ~ "Worst",
      TRUE ~ "Middle"
    ))
  ) %>%
  inner_join(statcast_clean, by=c("game_year", "pitcher")) %>%
  filter(pitch_type == "SL",tier %in% c("Best", "Worst"))

test %>%
  ggplot(aes(x=pfx_x*12, pfx_z*12, color=tier)) +
  stat_ellipse() +
  facet_wrap(~p_throws.y) +
  scale_x_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  scale_y_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  coord_equal()
```

Fastball speed by age:
```{r}
pitch_names <- c(
  "FF" = "4-Seam",
  "FC" = "Cutter",
  "FS" = "Splitter",
  "SI" = "2-Seam/Sinker"
)

fastballs <- statcast_clean %>%
  select(pitch_type, age_pit, release_speed, pitch_name) %>%
  filter(pitch_type == "CH"| pitch_type == "CU"| pitch_type == "FF"| pitch_type == "SI"| pitch_type == "SL"| pitch_type == "FC"| pitch_type == "ST"| pitch_type == "KC" | pitch_type == "FS") %>%
  mutate(pitch_type = recode(pitch_type,
                             "FF" = "4-Seam",
                             "SL" = "Slider",
                             "CH" = "Changeup",
                             "CU" = "Curveball",
                             "SI" = "2-Seam",
                             "KC"= "Knuckle Curve",
                             "FC" = "Cutter",
                             "ST" = "Sweeper",
                             "FS" = "Splitter")) %>%
  group_by(age_pit, pitch_type) %>%
  summarise(mean_speed = mean(release_speed, na.rm=TRUE))

ggplot(fastballs, aes(x=age_pit, 
                      y=mean_speed)) +
  geom_point(alpha = .5, 
             size = 2) + 
  geom_smooth(aes(color=pitch_type),
              se=FALSE) +
  facet_wrap(.~pitch_type, 
             nrow = 3) +
  force_panelsizes(rows = unit(5, "cm")) +
  labs(x = "Age", 
       y = "Average Pitch Speed (mph)", 
       color = "Pitch Type", 
       title = "Average Speed for Fastball Types at Each Recorded Age") + 
  theme(strip.text = element_blank(),
        strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1)
  ) 
```


Outcomes of each pitch data cleaning:
```{r}
pitch_type_outcomes <- statcast_clean %>%
  select(pitch_type, events) %>%
  filter(pitch_type == "CH"| pitch_type == "CU"| pitch_type == "FF"| pitch_type == "SI"| pitch_type == "SL"| pitch_type == "FC"| pitch_type == "ST"| pitch_type == "KC" | pitch_type == "FS") %>%
    filter(!is.na(events)) %>%
    mutate(outcome = case_when(
    events == "double" | events == "triple" | events == "home_run" ~ "XBH",
    events == "grounded_into_double_play" | events == "strikeout_double_play" | events == "double_play" | events == "sac_fly_double_play" | events == "triple_play" | events == "sac_bunt_double_play" ~ "Double/Triple Play",
    events == "field_out" | events == "force_out" | events == "fielders_choice" | events == "fielders_choice_out" ~ "Field Out",
    events == "strikeout" ~ "K",
    events == "single" ~ "Single",
    events == "walk" | events == "hit_by_pitch" ~ "Walk/HBP",
    events == "catcher_interf"  | events == "truncated_pa" | events == "sac_bunt" | events == "sac_fly" | events == "field_error" ~ "Other"
     )) %>%
  group_by(outcome, pitch_type) %>%
  summarize(count = n(), .groups = "drop_last") %>%
  mutate(prop = count / sum(count)) %>%
  mutate(pitch_type = recode(pitch_type,
                             "FF" = "4-Seam",
                             "SL" = "Slider",
                             "CH" = "Changeup",
                             "CU" = "Curveball",
                             "SI" = "2-Seam",
                             "KC"= "Knuckle Curve",
                             "FC" = "Cutter",
                             "ST" = "Sweeper",
                             "FS" = "Splitter")) %>%
  mutate(pitch_type = factor(pitch_type, levels = c(
    "4-Seam", "2-Seam", "Cutter", "Splitter", "Changeup", "Slider", "Sweeper", "Knuckle Curve", "Curveball"
  )))
```

Outcomes of each pitch visualization:
```{r}
ggplot(pitch_type_outcomes, aes(x="", y=prop, fill=pitch_type)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) + 
  facet_wrap(~outcome) +
  theme(strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1),
        axis.text = element_blank(),
        axis.ticks = element_blank()
  ) +
  labs(x = NULL, y = NULL, title = "Plate Appearance Outcomes by Pitch Type", fill = "PA Outcome")

counts <- statcast_clean %>%
  select(pitch_type) %>%
  filter(pitch_type == "CH"| pitch_type == "CU"| pitch_type == "FF"| pitch_type == "SI"| pitch_type == "SL"| pitch_type == "FC"| pitch_type == "ST"| pitch_type == "KC" | pitch_type == "FS") %>%
  group_by(pitch_type) %>%
  summarize(count=n()) %>%
  mutate(prop = count/sum(count)) %>%
  mutate(pitch_type = recode(pitch_type,
                             "FF" = "4-Seam",
                             "SL" = "Slider",
                             "CH" = "Changeup",
                             "CU" = "Curveball",
                             "SI" = "2-Seam",
                             "KC"= "Knuckle Curve",
                             "FC" = "Cutter",
                             "ST" = "Sweeper",
                             "FS" = "Splitter")) %>%
  mutate(pitch_type = factor(pitch_type, levels = c(
    "4-Seam", "2-Seam", "Cutter", "Splitter", "Changeup", "Slider", "Sweeper", "Knuckle Curve", "Curveball"
  )))

ggplot(counts, aes(x="", y=prop, fill=pitch_type)) +
  geom_bar(stat="identity", width=1, color="white") +
  geom_label(data = counts, aes(y=prop, x=1.6, label=scales::percent(prop, accuracy = 0.1), color = pitch_type),
             fill = "white",
             show.legend = FALSE,
             position = position_stack(vjust = .5)) +
  coord_polar("y", start=0) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1),
        strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = NULL, y = NULL)
```

