---
title: "Section B"
output: html_document
date: "2026-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries & Data

### Installing Libraries
```{r}
library(tidyverse)
library(knitr)
library(readr)
library(ggrepel)
library(sf)
library(scales)
library(ggh4x)
```

### Loading Data
```{r}
statcast <- read_csv("statcast_full.csv")
```

### Removing Unnecessary Columns
```{r}
statcast_clean <- statcast %>%
  select(-matches(".*deprecated$"), -spin_dir, -umpire) %>% #Remove deprecated numbers
  select(-matches("fielder_[0-9]")) %>% 
  select(-matches("on_[1-3]b$")) %>%
  select(-matches("[io]f_fielding_alignment")) %>%
  select(-matches("delta_.*")) %>%
  select(-matches("intercept_ball_.*")) %>%
  select(-matches(".*_days_since_.*"), -matches(".*_days_until_.*")) %>%
  select(-age_pit_legacy, -age_bat_legacy) %>%
  select(-matches("api_.*")) %>%
  select(-woba_value, -woba_denom) %>%
  select(-matches(".*score.*"))
  
statcast_clean <- statcast_clean %>%
  drop_na(pitch_type) %>%
  filter(game_type=="R")
```

### Formatting Columns
```{r}
statcast_clean <- statcast_clean %>%
  mutate(pitch_type = as.factor(pitch_type)) %>%
  mutate(description = as.factor(description)) %>%
  mutate(zone = as.factor(zone)) %>%
  mutate(p_throws = as.factor(p_throws)) %>%
  mutate(pitch_type = recode(pitch_type,
                             "FF" = "4-Seam",
                             "SI" = "2-Seam",
                             "FC" = "Cutter",
                             "FS" = "Splitter",
                             "CH" = "Changeup",
                             "KC"= "Knuckle Curve",
                             "CU" = "Curveball",
                             "SL" = "Slider",
                             "ST" = "Sweeper"
                             )) %>%
  mutate(pitch_type = factor(pitch_type, levels = c(
    "4-Seam", "2-Seam", "Cutter", "Splitter", "Changeup", "Knuckle Curve", "Curveball", "Slider", "Sweeper")))
```

### Assigning Colors to Pitches
```{r}
pitch_type_colors <- c(
  "4-Seam" = "#8B2323",
  "2-Seam" = "#FF3030",
  "Cutter" = "#CD6839",
  "Splitter" = "#FA8072",
  "Changeup" = "#97FFFF",
  "Knuckle Curve" = "#00B2EE",
  "Curveball" = "#4F94CD",
  "Slider" = "#3A5FCD",
  "Sweeper" = "#473C8B")
```

## Plots

### Pitch Mix
```{r}
pitch_props <- statcast_clean %>%
  group_by(game_year, pitch_name, pitch_type) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(game_year) %>%
  mutate(prop = n/sum(n)) %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             ))
pitch_props %>%
  ggplot(aes(x=game_year, y=prop, fill=pitch_type)) +
  geom_area(aes(fill=pitch_type)) +
  geom_label(
    aes(
      label = ifelse(
        pitch_type %in% c("Sweeper", "Slider", "2-Seam") & 
        game_year %in% c(2017, 2019, 2021, 2023, 2025),
        scales::percent(prop, accuracy = 0.1), 
        NA 
      ),
      color = pitch_type
    ),
    position = position_stack(vjust = 0.5),
    size = 3,
    fill = "white",
    show.legend = FALSE,
    na.rm = TRUE 
  ) +
  scale_fill_manual(values = pitch_type_colors) +
  scale_color_manual(values=pitch_type_colors) +
  theme_bw() +
  labs(title="Evolution of Pitch Mix",
       subtitle="Statcast Dataset",
       x="Season",
       y="Proportion",
       fill="Pitch Type") +
  scale_x_continuous(limits=c(2017,2025),
                     breaks=seq(2017,2025,1))
```

### Pitch Break
```{r}
centers <- statcast_clean %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%  
  group_by(pitch_type, pitch_name, p_throws) %>%
  summarise(
    x = mean(pfx_x, na.rm = TRUE)*12,
    y = mean(pfx_z, na.rm = TRUE)*12,
    .groups = "drop"
  )

statcast_clean %>%
  drop_na(pfx_x, pfx_z) %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%  
  mutate(pfx_x_in = 12 * pfx_x, pfx_z_in = 12 * pfx_z) %>%
  ggplot(aes(x=pfx_x_in, y=pfx_z_in, color=pitch_type)) +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  stat_ellipse(type="t", linetype=2, level=0.9) +
  geom_label_repel(data = centers,
            aes(x = x, y = y, label = pitch_type, color=pitch_type),
            size=2,
            show.legend = FALSE) +
  facet_wrap(~p_throws) +
  scale_x_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  scale_y_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  coord_equal() +
  theme_bw() +
  labs(title="Induced Movement of MLB Pitches by Pitch Type",
       subtitle = "Catcher's Perspective | Grouped by Pitcher Throwing Arm",
       x="Horizontal Movement",
       y="Induced Vertical Break",
       caption="Data Source: Statcast via baseballr",
       color="Pitch Type") +
  scale_color_manual(values = pitch_type_colors)
```

### Sweeper/Slider Movement Comparison
```{r}
centers_sl_st <- statcast_clean %>%
  filter(pitch_type %in% c("Slider", "Sweeper")) %>%
  group_by(pitch_type, pitch_name, p_throws) %>%
  summarise(
    x = mean(pfx_x, na.rm = TRUE)*12,
    y = mean(pfx_z, na.rm = TRUE)*12,
    .groups = "drop"
  )

statcast_clean %>%
  drop_na(pfx_x, pfx_z) %>%
  filter(pitch_type %in% c("Slider", "Sweeper")) %>%
  mutate(pfx_x_in = 12 * pfx_x, pfx_z_in = 12 * pfx_z) %>%
  ggplot(aes(x=pfx_x_in, y=pfx_z_in, color=pitch_type)) +
  geom_hline(yintercept=0) +
  geom_vline(xintercept=0) +
  stat_ellipse(type="t", linetype=2, level=0.9) +
  geom_label_repel(data = centers_sl_st,
            aes(x = x, y = y, label = pitch_type, color=pitch_type),
            size=2,
            show.legend = FALSE) +
  facet_wrap(~p_throws) +
  scale_x_continuous(limits=c(-25,25),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  scale_y_continuous(limits=c(-20,20),
                     breaks = seq(-25,25,5),
                     labels = scales::number_format(suffix = "\"")) +
  theme_bw() +
  labs(title="Induced Movement of Sliders and Sweepers",
       subtitle = "Catcher's Perspective | Grouped by Pitcher Throwing Arm",
       x="Horizontal Movement",
       y="Induced Vertical Break",
       caption="Data Source: Statcast via baseballr",
       color="Pitch Type") +
  scale_color_manual(values = pitch_type_colors)
```

### Sweeper/Slider Spin & Speed Comparison
```{r}
spin_and_speed <- statcast_clean %>%
  filter(pitch_type %in% c("Slider", "Sweeper")) %>%
  drop_na(release_speed, release_spin_rate) %>%
  group_by(pitch_type, game_year) %>%
  summarise(avg_spin = mean(release_spin_rate), avg_speed = mean(release_speed), .groups = "drop") %>%
  pivot_longer(cols = c(avg_speed, avg_spin), names_to="metric") %>%
  complete(pitch_type = c("Slider", "Sweeper"), metric, game_year = full_seq(game_year, 1))

mean_spin_and_speed <- spin_and_speed %>%
  group_by(metric, pitch_type) %>%
  summarise(mean_value = mean(value, na.rm=TRUE), .groups="drop")

spin_and_speed %>%
  ggplot(aes(x=game_year, y=value, color=pitch_type)) +
  geom_point() +
  geom_line() +
  geom_hline(data=mean_spin_and_speed,
            aes(yintercept=mean_value, color=pitch_type),
            linetype="dashed",
            linewidth = 1) +
  facet_wrap(~ metric, 
             scales = "free_y",
             labeller = labeller(
               metric = c(
                 avg_speed = "Average Release Speed (mph)",
                 avg_spin = "Average Spin Rate (rpm)"
               )
             )) +
  theme_bw() +
  labs(title="Spin Rate and Release Speed Comparison",
       subtitle = "Sliders vs Sweepers",
       caption="Data Source: Statcast via baseballr",
       x="Year",
       y="Speed/Spin Rate",
       color="Pitch Type") +
  scale_color_manual(values = pitch_type_colors)
  
```

### Pitch Speed by Age
```{r}
fastballs <- statcast_clean %>%
  select(pitch_type, age_pit, release_speed, pitch_name) %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%  
  group_by(age_pit, pitch_type) %>%
  summarise(mean_speed = mean(release_speed, na.rm=TRUE))

ggplot(fastballs, aes(x=age_pit, 
                      y=mean_speed,
                      color = pitch_type)) +
  geom_point(alpha = .5, 
             size = 2,
             color = "black") + 
  geom_smooth(se=FALSE) +
  facet_wrap(.~pitch_type, 
             nrow = 3) +
  scale_color_manual(values = pitch_type_colors) +
  force_panelsizes(rows = unit(5, "cm")) +
  labs(x = "Age", 
       y = "Average Pitch Speed (mph)", 
       color = "Pitch Type", 
       title = "Average Speed for Fastball Types at Each Recorded Age") + 
  theme(strip.text = element_blank(),
        strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1)
  )
```
This visualization shows the relationship between the age of pitchers and their pitch speeds for the 9 most common pitch types from 2017-2025. We can learn from this model that pitchers' speeds decline as they get older, which provides evidence for the assumptions that wear & tear as players age does occur. This may present questions regarding exact reasons as to why players' pitches slow down or speed up, such as injury or recovery from injury.

### PA Outcomes by Pitch Type Data Manipulation
```{r}
pitch_type_outcomes <- statcast_clean %>%
  select(pitch_type, events) %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%      filter(!is.na(events)) %>%
    mutate(outcome = case_when(
    events == "double" | events == "triple" | events == "home_run" ~ "XBH",
    events == "grounded_into_double_play" | events == "strikeout_double_play" | events == "double_play" | events == "sac_fly_double_play" | events == "triple_play" | events == "sac_bunt_double_play" ~ "Double/Triple Play",
    events == "field_out" | events == "force_out" | events == "fielders_choice" | events == "fielders_choice_out" ~ "Field Out",
    events == "strikeout" ~ "K",
    events == "single" ~ "Single",
    events == "walk" | events == "hit_by_pitch" ~ "Walk/HBP",
    events == "catcher_interf"  | events == "truncated_pa" | events == "sac_bunt" | events == "sac_fly" | events == "field_error" ~ "Other"
     )) %>%
  mutate(pitch_type = factor(pitch_type, levels = c(
    "4-Seam", "2-Seam", "Cutter", "Splitter", "Changeup", "Knuckle Curve", "Curveball", "Slider", "Sweeper"))) %>%
  group_by(outcome, pitch_type) %>%
  summarize(count = n(), .groups = "drop_last") %>%
  mutate(prop = count / sum(count))
```

### PA Outcomes by Pitch Type Visualizations
```{r}
ggplot(pitch_type_outcomes, aes(x="", y=prop, fill=pitch_type)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) + 
  facet_wrap(~outcome) +
  scale_fill_manual(values = pitch_type_colors) +
  theme(strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1),
        axis.text = element_blank(),
        axis.ticks = element_blank()
  ) +
  labs(x = NULL, y = NULL, title = "Plate Appearance Outcomes by Pitch Type", fill = "PA Outcome")

counts <- statcast_clean %>%
  select(pitch_type) %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%    group_by(pitch_type) %>%
  summarize(count=n()) %>%
  mutate(prop = count/sum(count))

ggplot(counts, aes(x="", y=prop, fill=pitch_type)) +
  geom_bar(stat="identity", width=1, color="white") +
  geom_label(data = counts, aes(y=prop, x=1.6, label=scales::percent(prop, accuracy = 0.1), color = pitch_type),
             fill = "white",
             show.legend = FALSE,
             position = position_stack(vjust = .5)) +
  coord_polar("y", start=0) +
  scale_fill_manual(values = pitch_type_colors) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1),
        strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = NULL, y = NULL)
```
This visualization shows the percentage of each pitch that makes up each type of plate appearance and we can understand from this pie chart that pitches categorized as a form of fastball are typically the most common in several categories, such as double plays, walks, and singles. This may lead to research into why certain pitches are more hittable than others; this may have several causes, such as pitch accuracy or speed.

### Specific Counts vs. Pitch Types
```{r}
specific_counts <- statcast_clean %>%
  select(pitch_type, balls, strikes) %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%    mutate(batting_count = paste(balls, strikes, sep = "-")) %>%
  group_by(batting_count, pitch_type) %>%
  summarize(count = n()) %>%
  mutate(prop = count/sum(count))

ggplot(specific_counts, aes(x="", y=prop, fill=pitch_type)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  facet_wrap(~batting_count, nrow=4) +
  scale_fill_manual(values = pitch_type_colors) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1),
        strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = NULL, y = NULL)

```
This visualization shows a pie chart for every single possible count in baseball, split by the proportion of each pitch thrown. This visualization shows that fastballs are very common on counts with either 3 balls or 0 strikes, while a breaking ball, such as a slider, is relatively more common in 1-2 or 2-2 counts. This visualization may lead to further questioning into pitch sequencing throughout an entire plate appearance.

### Specific Counts vs. Outcomes
```{r}
specific_counts_outcomes <- statcast_clean %>%
  select(description, balls, strikes) %>%
  mutate(batting_count = paste(balls, strikes, sep = "-")) %>%
  filter(!is.na(description)) %>%
    mutate(outcome = case_when(
      description == "hit_into_play" ~ "Hit",
      description == "swinging_strike" | description == "swinging_pitchout" | description == "missed_bunt" ~ "Swinging K",
      description == "called_strike" ~ "Called Strike",
      description == "foul" | description == "foul_bunt" | description == "bunt_foul_tip" | description == "foul_tip" | description == "foul_pitchout" ~ "Foul",
      description == "pitchout" ~ "Pitchout",
      description == "ball" | description == "blocked_ball" | description == "hit_by_pitch" ~ "Ball/HBP",
      description == "swinging_strike_blocked" ~ "Dropped Swinging Strike")) %>%
  group_by(batting_count, outcome) %>%
  summarize(count = n()) %>%
  mutate(prop = count/sum(count))

ggplot(specific_counts_outcomes, aes(x="", y=prop, fill=outcome)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  facet_wrap(~batting_count, nrow=4) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1),
        strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = NULL, y = NULL)

```
This visualization shows the outcome of each pitch during every possible count, which helps us understand that batters are much more likely to get a hit on counts with 2 strikes rather than counts with 0 strikes, and are very likely to take on 3-0 counts. This pattern of behavior can lead to questions relating to patterns about which pitches are usually hit or swung on in specific counts.

### Pitch Type vs. Outcome
```{r}
pitch_outcomes_with_pitch_types <- statcast_clean %>%
  select(description, pitch_type) %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%
  filter(!is.na(description)) %>%
    mutate(outcome = case_when(
      description == "hit_into_play" ~ "Hit",
      description == "swinging_strike" | description == "swinging_pitchout" | description == "missed_bunt" ~ "Swinging K",
      description == "called_strike" ~ "Called Strike",
      description == "foul" | description == "foul_bunt" | description == "bunt_foul_tip" | description == "foul_tip" | description == "foul_pitchout" ~ "Foul",
      description == "pitchout" ~ "Pitchout",
      description == "ball" | description == "blocked_ball" | description == "hit_by_pitch" ~ "Ball/HBP",
      description == "swinging_strike_blocked" ~ "Dropped Swinging Strike")) %>%
  group_by(pitch_type, outcome) %>%
  mutate(pitch_type = factor(pitch_type, levels = c(
    "4-Seam", "2-Seam", "Cutter", "Splitter", "Changeup", "Knuckle Curve", "Curveball", "Slider", "Sweeper"))) %>%
  summarize(count = n()) %>%
  mutate(prop = count/sum(count))

ggplot(pitch_outcomes_with_pitch_types, aes(x="", y=prop, fill=outcome)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  facet_wrap(~pitch_type, nrow=3) +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "black",
                                    fill = NA,
                                    linewidth = 1),
        strip.background = element_blank(),
        panel.background= element_rect(fill="gray89"),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  labs(x = NULL, y = NULL)
```
This visualization showed the outcome proportions of each different type of pitch, which shows that, overall, pitch outcomes are quite similar for each type of pitch. This would lead to further questioning into the efficacy of different pitches at different points in the count.

### RHP Relative Zone Location by Pitch Type
```{r}
zone1coords <- matrix(c(-17/2,17/2, -17/6,17/2, -17/6,17/6, -17/2,17/6, -17/2,17/2),
                      ncol = 2,
                      byrow= TRUE)
zone2coords <- matrix(c(-17/6,17/2, 17/6,17/2, 17/6,17/6, -17/6,17/6, -17/6,17/2),
                      ncol = 2,
                      byrow=TRUE)
zone3coords <- matrix(c(17/6,17/2, 17/2,17/2, 17/2,17/6, 17/6,17/6, 17/6,17/2),
                      ncol = 2,
                      byrow=TRUE)
zone4coords <- matrix(c(-17/2,17/6, -17/6,17/6, -17/6,-17/6, -17/2,-17/6, -17/2,17/6),
                      ncol = 2,
                      byrow=TRUE)
zone5coords <- matrix(c(-17/6,17/6, 17/6,17/6, 17/6,-17/6, -17/6,-17/6, -17/6,17/6),
                      ncol = 2,
                      byrow=TRUE)
zone6coords <- matrix(c(17/6,17/6, 17/2,17/6, 17/2,-17/6, 17/6,-17/6, 17/6,17/6),
                      ncol = 2,
                      byrow=TRUE)
zone7coords <- matrix(c(-17/2,-17/6, -17/6,-17/6, -17/6,-17/2, -17/2,-17/2, -17/2,-17/6),
                      ncol = 2,
                      byrow= TRUE)
zone8coords <- matrix(c(-17/6,-17/6, 17/6,-17/6, 17/6,-17/2, -17/6,-17/2, -17/6,-17/6),
                      ncol = 2,
                      byrow=TRUE)
zone9coords <- matrix(c(17/6,-17/6, 17/2,-17/6, 17/2,-17/2, 17/6,-17/2, 17/6,-17/6),
                      ncol = 2,
                      byrow=TRUE)
zone11coords <- matrix(c(0,17/2, -17/2,17/2, -17/2,0, -12.3,0, -12.3,12.3, 0,12.3, 0,17/2),
                       ncol=2,
                       byrow=TRUE)
zone12coords <- matrix(c(0,17/2, 17/2,17/2, 17/2,0, 12.3,0, 12.3,12.3, 0,12.3, 0,17/2),
                       ncol=2,
                       byrow=TRUE)
zone13coords <- matrix(c(17/2,0, 12.3,0, 12.3,-12.3, 0,-12.3, 0,-17/2, 17/2,-17/2, 17/2, 0),
                       ncol=2,
                       byrow=TRUE)
zone14coords <- matrix(c(0,-17/2, 0,-12.3, -12.3,-12.3, -12.3,0, -17/2,0, -17/2,-17/2, 0,-17/2),
                       ncol=2,
                       byrow=TRUE)

zones <- st_sfc(st_polygon(list(zone1coords)), 
                st_polygon(list(zone2coords)),
                st_polygon(list(zone3coords)),
                st_polygon(list(zone4coords)),
                st_polygon(list(zone5coords)),
                st_polygon(list(zone6coords)),
                st_polygon(list(zone7coords)),
                st_polygon(list(zone8coords)),
                st_polygon(list(zone9coords)),
                st_polygon(list(zone11coords)),
                st_polygon(list(zone12coords)),
                st_polygon(list(zone13coords)),
                st_polygon(list(zone14coords)))
areas <- st_sf(
  Strike_Zone = c("Zone 1", "Zone 2", "Zone 3", "Zone 4", "Zone 5", "Zone 6", "Zone 7", "Zone 8", "Zone 9", "Zone 11", "Zone 12", "Zone 13", "Zone 14"),
  geometry = zones
)

strike_zone_data <- statcast_clean %>%
  select(pitch_type, zone, p_throws) %>%
  filter(p_throws == "R") %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%
  mutate(pitch_type = factor(pitch_type, levels = c(
    "4-Seam", "2-Seam", "Cutter", "Splitter", "Changeup", "Knuckle Curve", "Curveball", "Slider", "Sweeper"))) %>%
  filter(!is.na(zone)) %>%
  mutate(Strike_Zone = recode(zone,
                       "1" = "Zone 1",
                       "2" = "Zone 2",
                       "3" = "Zone 3",
                       "4" = "Zone 4",
                       "5" = "Zone 5",
                       "6" = "Zone 6",
                       "7" = "Zone 7",
                       "8" = "Zone 8",
                       "9" = "Zone 9",
                       "11" = "Zone 11",
                       "12" = "Zone 12",
                       "13" = "Zone 13",
                       "14" = "Zone 14")) %>%
  group_by(pitch_type, Strike_Zone) %>%
  summarize(count=n()) %>%
  mutate(percent = count/sum(count)*100)

print(areas)
plot(areas["Strike_Zone"], col = NA)

areas_with_data <- areas %>%
  left_join(strike_zone_data, by = "Strike_Zone") %>%
  group_by(pitch_type, Strike_Zone)

ggplot(areas_with_data) + 
  geom_sf(aes(fill=percent), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "mako", name = "Pitch %", na.value = "grey90", direction = -1) +
  facet_wrap(~pitch_type) +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank())
```

### LHP Relative Zone Location by Pitch Type
```{r}
zone1coords <- matrix(c(-17/2,17/2, -17/6,17/2, -17/6,17/6, -17/2,17/6, -17/2,17/2),
                      ncol = 2,
                      byrow= TRUE)
zone2coords <- matrix(c(-17/6,17/2, 17/6,17/2, 17/6,17/6, -17/6,17/6, -17/6,17/2),
                      ncol = 2,
                      byrow=TRUE)
zone3coords <- matrix(c(17/6,17/2, 17/2,17/2, 17/2,17/6, 17/6,17/6, 17/6,17/2),
                      ncol = 2,
                      byrow=TRUE)
zone4coords <- matrix(c(-17/2,17/6, -17/6,17/6, -17/6,-17/6, -17/2,-17/6, -17/2,17/6),
                      ncol = 2,
                      byrow=TRUE)
zone5coords <- matrix(c(-17/6,17/6, 17/6,17/6, 17/6,-17/6, -17/6,-17/6, -17/6,17/6),
                      ncol = 2,
                      byrow=TRUE)
zone6coords <- matrix(c(17/6,17/6, 17/2,17/6, 17/2,-17/6, 17/6,-17/6, 17/6,17/6),
                      ncol = 2,
                      byrow=TRUE)
zone7coords <- matrix(c(-17/2,-17/6, -17/6,-17/6, -17/6,-17/2, -17/2,-17/2, -17/2,-17/6),
                      ncol = 2,
                      byrow= TRUE)
zone8coords <- matrix(c(-17/6,-17/6, 17/6,-17/6, 17/6,-17/2, -17/6,-17/2, -17/6,-17/6),
                      ncol = 2,
                      byrow=TRUE)
zone9coords <- matrix(c(17/6,-17/6, 17/2,-17/6, 17/2,-17/2, 17/6,-17/2, 17/6,-17/6),
                      ncol = 2,
                      byrow=TRUE)
zone11coords <- matrix(c(0,17/2, -17/2,17/2, -17/2,0, -12.3,0, -12.3,12.3, 0,12.3, 0,17/2),
                       ncol=2,
                       byrow=TRUE)
zone12coords <- matrix(c(0,17/2, 17/2,17/2, 17/2,0, 12.3,0, 12.3,12.3, 0,12.3, 0,17/2),
                       ncol=2,
                       byrow=TRUE)
zone13coords <- matrix(c(17/2,0, 12.3,0, 12.3,-12.3, 0,-12.3, 0,-17/2, 17/2,-17/2, 17/2, 0),
                       ncol=2,
                       byrow=TRUE)
zone14coords <- matrix(c(0,-17/2, 0,-12.3, -12.3,-12.3, -12.3,0, -17/2,0, -17/2,-17/2, 0,-17/2),
                       ncol=2,
                       byrow=TRUE)

zones <- st_sfc(st_polygon(list(zone1coords)), 
                st_polygon(list(zone2coords)),
                st_polygon(list(zone3coords)),
                st_polygon(list(zone4coords)),
                st_polygon(list(zone5coords)),
                st_polygon(list(zone6coords)),
                st_polygon(list(zone7coords)),
                st_polygon(list(zone8coords)),
                st_polygon(list(zone9coords)),
                st_polygon(list(zone11coords)),
                st_polygon(list(zone12coords)),
                st_polygon(list(zone13coords)),
                st_polygon(list(zone14coords)))
areas <- st_sf(
  Strike_Zone = c("Zone 1", "Zone 2", "Zone 3", "Zone 4", "Zone 5", "Zone 6", "Zone 7", "Zone 8", "Zone 9", "Zone 11", "Zone 12", "Zone 13", "Zone 14"),
  geometry = zones
)

strike_zone_data <- statcast_clean %>%
  select(pitch_type, zone, p_throws) %>%
  filter(p_throws == "L") %>%
  filter(pitch_type %in% c("4-Seam",
                             "2-Seam",
                             "Cutter",
                             "Splitter",
                             "Changeup",
                             "Knuckle Curve",
                             "Curveball",
                             "Slider",
                             "Sweeper"
                             )) %>%
  mutate(pitch_type = factor(pitch_type, levels = c(
    "4-Seam", "2-Seam", "Cutter", "Splitter", "Changeup", "Knuckle Curve", "Curveball", "Slider", "Sweeper"))) %>%
  filter(!is.na(zone)) %>%
  mutate(Strike_Zone = recode(zone,
                       "1" = "Zone 1",
                       "2" = "Zone 2",
                       "3" = "Zone 3",
                       "4" = "Zone 4",
                       "5" = "Zone 5",
                       "6" = "Zone 6",
                       "7" = "Zone 7",
                       "8" = "Zone 8",
                       "9" = "Zone 9",
                       "11" = "Zone 11",
                       "12" = "Zone 12",
                       "13" = "Zone 13",
                       "14" = "Zone 14")) %>%
  group_by(pitch_type, Strike_Zone) %>%
  summarize(count=n()) %>%
  mutate(percent = count/sum(count)*100)

print(areas)
plot(areas["Strike_Zone"], col = NA)

areas_with_data <- areas %>%
  left_join(strike_zone_data, by = "Strike_Zone") %>%
  group_by(pitch_type, Strike_Zone)

ggplot(areas_with_data) + 
  geom_sf(aes(fill=percent), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "mako", name = "Pitch %", na.value = "grey90", direction = -1) +
  facet_wrap(~pitch_type) +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank())
```
Although the RHP and LHP visualizations are separate, they serve as one visualization simply showing two plots differentiated by handedness. We can learn from this that more balls are likely to be thrown when a pitcher is throwing a breaking ball, which can be connected to the article by Sport Journal which investigates the relationship between pitch accuracy and batter outcomes as certain pitches that are balls may turn into swinging strikes due to the movement of the pitch, which further prompts questions regarding the percentage of would-be balls that are swung at.